An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-02-07 latex_files/ch5/5-1-ma2.tex
addressed henri's comments

\begin{table}
\begin{center}
\begin{tabular}{ c|c|c|c|c }
\hline
& \(\mu_{\theta_1}\) & \(\sigma_{\theta_1}\) & \(\mu_{\theta_2}\) & \(\sigma_{\theta_2}\) \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.503 & 0.143 & 0.032 & 0.17 \\
\hline
ROMC (Bayesian optimisation) & 0.494 & 0.16 & 0.086 & 0.167 \\
\hline
ROMC (Neural Network) & 0.503 & 0.140 & 0.03 & 0.171 \\
\hline
\end{tabular}
\end{center}
\caption{\textbf{HP comment: Needs a fine caption to get Table number.}\label{tab:ma2}}
\end{table}
% \begin{figure}[H]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist.png}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/ma2_distance_hist_bo.png} \\
%     \end{center}
%     \caption[MA2 example, histogram of distances]{Histogram of distances
%       \(d_i^*, i \in \{ 1, \ldots, n_1 \}\). The left graph corresponds
%       to the gradient-based approach and the right one to the Bayesian
%       optimisation approach.}
%   \label{fig:ma2_2}
% \end{figure}

@@

\begin{figure}[ht]
  \centering
    \begin{minipage}[t]{.49\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Solve optimisation problems},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.49\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Proposal regions},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.49\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Sample},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.49\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Evaluate posterior},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \caption[Execution time exploiting parallelisation]{In the first
      line, we compare the parallel and sequential execution of the
      training part and at the second line the inference part. At the
      left figure, we measure the execution time for sampling \(n_2=50\)
      points per region. At the right figure we measure the execution
      time for evaluating the posterior at a batch of \(50\) points.}
      \label{fig:exec_parallel}
\end{figure}

>>>>

\begin{table}
\begin{center}
\begin{tabular}{ c|c|c|c|c }
\hline
& \(\mu_{\theta_1}\) & \(\sigma_{\theta_1}\) & \(\mu_{\theta_2}\) & \(\sigma_{\theta_2}\) \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.503 & 0.143 & 0.032 & 0.17 \\
\hline
ROMC (Bayesian optimisation) & 0.494 & 0.16 & 0.086 & 0.167 \\
\hline
ROMC (Neural Network) & 0.503 & 0.140 & 0.03 & 0.171 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the samples obtained from the estimated
  posterior with (a) Rejection sampling and (b) the different versions
  of ROMC. We observe that the obtained samples share similar
  statistics along all methods. \label{tab:ma2}}
\end{table}

@@

\begin{figure}[ht]
  \centering
    \begin{minipage}[t]{.4\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Solve optimisation problems},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.4\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Proposal regions},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.4\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Sample},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}[t]{.4\textwidth}
      \begin{tikzpicture}
        \begin{axis}[
          title={Evaluate posterior},
          xlabel={\(n_1\)},
          ylabel={seconds},
          xmin=1, xmax=500,
          ymin=0, ymax=7,
          xtick={1,100,200,300,400,500},
          ytick={0, 1, 2, 3, 4, 5, 6, 7},
          legend pos=north west,
          ymajorgrids=true,
          grid style=dashed,
          ]
          \addplot[color=blue, mark=o] coordinates {
            (1,0.05)(125,1.6)(250,3.5)(375,5)(500,6.8)
          };
          \addlegendentry{sequential}
          \addplot[color=red, mark=o] coordinates {
            (1,0.2)(125,0.6)(250,1)(375,1.5)(500,2)
          };
          \addlegendentry{parallel}          
        \end{axis}
      \end{tikzpicture}
    \end{minipage}\hfill
    \caption[Execution time exploiting parallelisation]{In the first
      line, we compare the parallel and sequential execution of the
      training part and at the second line the inference part. At the
      left figure, we measure the execution time for sampling \(n_2=50\)
      points per region. At the right figure we measure the execution
      time for evaluating the posterior at a batch of \(50\) points.}
      \label{fig:exec_parallel}
\end{figure}

### 2022-02-07 paper.tex
addressed henri's comments

%% - \Abstract{} almost as usual
\Abstract{ Performing the inference in statistical models with
  intractable likelihood is challenging; most likelihood-free
  inference (LFI) methods encounter problems in accuracy and
  efficiency. In this paper, we present the implementation of Robust
  Optimisation Monte Carlo (ROMC) in the Python package ELFI. ROMC is
  a novel and efficient (highly-parallelizable) LFI technique that
  provides accurate weighted samples from the posterior.  Our
  implementation can be used in two ways. A scientist may use it as an
  out-of-the-box LFI algorithm. We provide an easy-to-use API
  harmonised with the rest of the ELFI framework; therefore, it is
  straightforward to compare it with all the methods included in
  ELFI. On the other direction, we have carefully split ROMC into
  isolated algorithmic steps. Therefore, we allow the researcher to
  experiment, replacing components of the algorithm with
  custom-designed subparts. In both scenarios, the algorithmic steps
  of ROMC run in a fully-parallelizable manner, exploiting all CPU
  cores. We also provide helpful functionalities for (i) inspecting
  the inference process and (ii) evaluating the approximated
  posterior. Finally, we test the robustness of our implementation on
  some typical LFI examples.

  % This short article illustrates how to write a manuscript for the
  % \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
  % Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
  % it is recommended to keep the {\LaTeX} code as simple as possible,
  % i.e., avoid inclusion of packages/commands that are not necessary.
  % For outlining the typical structure of a JSS article some brief text snippets
  % are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
  % discussing count data regression in \proglang{R}. Editorial comments and
  % instructions are marked by vertical bars.
}

@@

Optimization Monte Carlo (OMC) proposed by~\citep{Meeds2015} was a
novel LFI approach for approximating the posterior
distribution. Afterwards, \citep{Forneron2016} provided a similar
method under `reverse sampler'. The key idea is turning the stochastic
data-generating mechanism into a set of fully-parallelizable
deterministic optimisation processes. In their work,
~\citet{Ikonomov2019}, located some critical failure modes of OMC, and
proposed Robust OMC (ROMC) an alternative version of OMC with the
appropriate improvements.

In this paper, we present an extendible and parallelisable
implementation of ROMC at the \proglang{Python} package
\proglang{Engine for Likelihood-Free inference (ELFI)}. We follow
specific principles for making the implementation extendible. As we
will describe analytically in the next chapter, ROMC can be understood
not only as a specific algorithm, but mainly as a framework for
obtaining weighted samples from the posterior; it defines some
algorithmic task that must be fulfilled, without enforcing which
specific algorithm must be used for the tasks. Therefore, a researcher
may adopt ROMC's backbone approach, but experiment with alternative
methods to solve the distinct tasks. We have designed ROMC for
facilitating such alterations. Finally, we have tested the accuracy and
the efficiency of our implementation on the LFI examples supported by
the \proglang{ELFI} package.

@@

% \clearpage
\section{Background}
We first give a short introduction to implicit parametric models and
LFI. We then focus on the methods relevant for
this paper: first the earlier OMC and then
its recent robust generalisation, robust OMC, for which we provide the
software package as part of ELFI, a larger inference platform that we briefly introduce at the
end of this background section.

@@

\subsection{Implicit models and likelihood-free inference}
\input{latex_files/ch2/2-1-implicit-models.tex}

>>>>

%% - \Abstract{} almost as usual
\Abstract{ Performing the inference in statistical models with
  intractable likelihood is challenging; most likelihood-free
  inference (LFI) methods encounter problems in accuracy and
  efficiency. In this paper, we present the implementation of Robust
  Optimisation Monte Carlo (ROMC) in the \proglang{Python} package
  \pkg{ELFI}. ROMC is a novel and efficient (highly-parallelizable)
  LFI framework that provides accurate weighted samples from the
  posterior. Our implementation can be used in two ways. Firstly, a
  scientist may use it as an out-of-the-box LFI algorithm; we provide
  an easy-to-use API harmonised with the rest of the \pkg{ELFI}
  framework.  Therefore, it is straightforward to compare it with the
  rest of the methods included in \pkg{ELFI}. Afterwards, we have
  carefully split ROMC into isolated algorithmic steps. Hence, we
  allow the researcher to experiment by replacing part(s) of the
  algorithm with novel subcomponent(s). In both scenarios, all steps
  of ROMC run in a fully-parallelizable manner, exploiting all CPU
  cores. We also provide helpful functionalities for (i) inspecting
  the inference process and (ii) evaluating the obtained
  samples. Finally, we test the robustness of our implementation on
  some typical LFI examples.  }

@@

Optimization Monte Carlo (OMC) proposed by~\citet{Meeds2015} was a
novel LFI approach for approximating the posterior
distribution. Afterwards, \citet{Forneron2016} provided a similar
method naming it `reverse sampler'. The central idea is turning the
stochastic data-generating mechanism into a set of
fully-parallelizable deterministic optimisation processes. In their
work, ~\citet{Ikonomov2019}, located some critical failure modes of
OMC, and proposed Robust OMC (ROMC) an alternative version of OMC with
the appropriate improvements.

In this paper, we present an extendible and parallelisable
implementation of ROMC at the \proglang{Python} package \pkg{Engine
  for Likelihood-Free inference (ELFI)}. We choose carefully the
designing principles for making the implementation extendible. As we
describe analytically in the next chapter, ROMC can be understood not
only as a specific algorithm, but mainly as a framework for obtaining
weighted samples from the posterior; it defines a sequence of
algorithmic steps, without enforcing a specific algorithm for solving
them. Therefore, a researcher may adopt ROMC's backbone approach, but
experiment with alternative methods to solve the tasks. We have
designed ROMC for facilitating such alterations. Finally, we have
tested the accuracy and the efficiency of our implementation on the
LFI examples supported by the \pkg{ELFI} package.

@@

% \clearpage
\section{Background}
We first give a short introduction to simulator-based models and
LFI. We then focus on the methods forming the basis of this paper; OMC
and its robust improvement, ROMC. Finally, we introduce ELFI, the
software package used for our implementation.

@@

\subsection{Simulator-based models and likelihood-free inference}
\input{latex_files/ch2/2-1-implicit-models.tex}
